---
title: "Data Visualization Fieldguide for Education"
author: "Jared E. Knowles"
date: "March 2015"
output: 
  rmarkdown::tufte_handout:
    keep_tex: true
---

```{r knitrsetup,echo=FALSE,eval=TRUE,results="hide",error=FALSE,message=FALSE,warning=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, echo=FALSE, comment=NA)
library(eeptools)
library(ggplot2)
library(ggthemes)
# source("data/simulate_data.R")
# load("data/cache.rda")
load("data/smalldata.rda")
```

\noindent \emph{Tufte's Tips}

- Have a properly chosen format and design
- Use words, numbers and drawing together
- Reflect balance, proportion, and a sense of relevant scale
- Display an accessible complexity of detail 
- Have a anarrative quality, tell a story about the data
```{r introplot, fig.margin = TRUE, fig.cap = "Sometimes the story is very simple."}
library(plyr)
plotdf <- ddply(iris, .(Species), summarize, Plength = mean(Petal.Length))
qplot(Species, Plength, data = plotdf, width = I(.5),
       geom='bar', 
      stat = 'identity') + theme_tufte() + 
  labs(y = "Petal Length (cm)", title = "Orchid Petal Size") 
```
- Draw in a professional manner, with the technical details done with care
- Avoid content-free decoration, including **chartjunk**

\noindent \emph{Planning}

\noindent Data visualization is a tool for communicating **a 
specific feature** of a dataset in an approachable and efficient manner.\marginnote{
Data visualization is done in service of the audience. If the goal is not to convey 
information, it is art!}

- Who is the audience? What is their background? Their biases?
- What is the context? What would the user consider to be *good news*, what would 
be *bad news*? How different are these? 
- What does the design style say? 
```{r badexcel, fig.margin = TRUE, fig.cap = "Theme says a lot. This says: `I am in a hurry'."}
qplot(Sepal.Length, Petal.Length, data = iris, color = Species) + theme_excel()
```
- Are you designing this visualization for a specific data set, or to display 
different data in a dashboard environment? 
- How long will this visualization last? Will it be updated next year? Does it need to be interactive?
- What visual elements best map onto the data I have -- shapes, sizes, colors, or fill?   \marginnote{How you turn dimensions in the into visual cues for your audience is everything.}
  
```{r diamondplot, fig.width = 8, fig.height = 2.1, fig.cap="Shape and color work well for discrete data without too many levels, but one may work better than the other. Sometimes a line of best fit works best.", fig.fullwidth = TRUE}
library(gridExtra)
data(diamonds)
diamonds <- diamonds[diamonds$cut != "Fair", ]
plotdf <- diamonds[sample(row.names(diamonds), 500), ]
plotdf$cut3 <- 0
plotdf$cut3 <- ifelse(plotdf$cut == "Ideal", 4, plotdf$cut3)
plotdf$cut3 <- ifelse(plotdf$cut == "Premium", 3, plotdf$cut3)
plotdf$cut3 <- ifelse(plotdf$cut == "Very Good", 2, plotdf$cut3)
plotdf$cut3 <- ifelse(plotdf$cut == "Good", 1, plotdf$cut3)
plotdf$cut2 <- factor(as.character(plotdf$cut))

p1 <- qplot(log(carat), log(price), data = plotdf, color = cut3) + 
  theme_bw(base_size = 8) +scale_color_continuous_tableau() + 
  guides(color= guide_legend(ncol = 2)) + 
  theme(legend.position = c(0.8, 0.35)) + 
  labs(x = "Log(carat)", y = "Log(price)", color = "Cut")+ 
  theme(legend.key.size = unit(3.5, "mm"), legend.text = element_text(size = 4),
        legend.key.width = unit(3, "mm"), legend.margin = unit(1, "mm"))

p2 <- qplot(log(carat), log(price), data = plotdf, shape = cut, alpha = I(0.4)) + 
  theme_bw(base_size = 8) + theme(legend.position = c(0.22, 0.78)) + 
  guides(shape= guide_legend(ncol = 2)) + 
  labs(x = "Log(carat)", y = "Log(price)", shape = "Cut") + 
  theme(legend.key.size = unit(4.5, "mm"), legend.text = element_text(size = 4),
        legend.key.width = unit(3, "mm"), legend.margin = unit(0, "mm"))

p3 <- qplot(log(carat), log(price), data = plotdf, alpha = I(0.4), 
            geom = 'smooth', se = FALSE, size = I(1.1), color = I("black"),
            linetype = cut) + 
  theme_bw(base_size = 8) + theme(legend.position = c(0.22, 0.78)) + 
  guides(linetype= guide_legend(ncol = 2)) + 
  labs(x = "Log(carat)", y = "Log(price)", linetype = "Cut")+ 
  theme(legend.key.size = unit(.45, "cm"), legend.text = element_text(size = 4),
        legend.key.width = unit(.25, "cm"), legend.margin = unit(0, "cm"))

grid.arrange(p1, p2, p3, ncol = 3, main = "Diamond Size and Price by Color and Cut")
```


```{r widesmallmultiples, eval=FALSE, fig.fullwidth = TRUE, fig.width = 10, fig.height = 2, fig.cap = "Small multiples spread the data out. But be careful, can the user draw meaningful distinctions between the groups from the data alone?"}
qplot(attday, readSS, data = df) + theme_bw() + 
  facet_wrap(~race, nrow=1) + labs(x = "Attendance Days", y = "Reading Score")
```



\noindent \emph{Dimensionality}

*If you got big data, I feel sorry for you son. you got $99^{n}$ problems but 
data viz ain't one.* [^jayz]
 
```{r bigplot, fig.fullwidth = TRUE, fig.cap = "Add reference lines and annotate your charts with text. Guide the viewer, do not assume. Craete visual context to invite and aid comparisons for the user.", fig.width = 10, fig.height = 2}
qplot(attday, readSS, data = df[df$race != "A", ], 
      alpha = I(0.5)) +   theme_bw() + 
  geom_hline(yintercept = 600, size = I(1.02), color = 'dark green') + 
  geom_smooth(method = 'lm') + 
#   geom_hline(yintercept = 530, , size = I(1.02), 
#              linetype = 2, color = I("dark blue")) +
#   geom_hline(yintercept = 480, , size = I(1.02), 
#              linetype = 2, color = I("goldenrod")) +
  geom_hline(yintercept = 420, , size = I(1.02), 
             linetype = 2, color = I("red")) + 
  annotate(geom="text", x = 140, y = 640, label = "Advanced", size = 3, 
           fontface = "bold") + 
#   annotate(geom="text", x = 135, y = 560, label = "Proficient", size = 3) + 
#   annotate(geom="text", x = 135, y = 510, label = "Basic", size = 3) + 
  annotate(geom="text", x = 140, y = 460, label = "Minimal", size = 3, 
           fontface = "bold") + 
      facet_wrap(~race, nrow = 1) + 
  labs(x = "Attendance Days", y = "Reading Scale Score") 
```


\noindent \emph{Review}

\noindent \textbf{Before you submit}, check that you have:

- **Axis labels** and a **title**: Is your purpose clear?
- A **legend**: Are all symbols explained?
- A **scale**: Is the relation of data points clear? 
- **Annotations**: Is there enough context? Are key points highlighted? 
\marginnote{Don't be afraid to model the data. A line of best fit or a smoother can provide a good summary to viewers to make comparisons across panels.}

\noindent Remember: If a picture is worth a thousand words, a good data visualization must always be *better than a table.*  
   
\noindent Review this with *fresh eyes*. Can someone who has not seen this data before undertand and interpret the main story here?   \marginnote{If it is not clear, consider encoding a key feature twice -- e.g. using color and shape.}
  
  
\noindent \emph{Technical Details}

\noindent The final format and use case should inform the design:

```{r technologies,echo=FALSE,fig.margin = TRUE, fig.cap = "An opinonated ordering of the tools available to do data viz. Not comprehensive or complete and your mileage will vary depending on your comfort with certain technologies."}
techs<-c("Adobe", "R", "Stata/SPSS", "D3", "Tableau", "JavaScript", 
         "Processing", "Excel", "SAS")
speed  <- c(1.5, 7, 5, 6, 9, 4, 5, 9, 4)
quality <- c(10, 9, 5, 9, 8, 8, 9, 2, 4)
web <- c("Yes", "Yes", "No", "Yes", "No", "Yes", "Yes", "No", "No")
viz <- data.frame(techs=techs, speed=speed, quality=quality, web=web)

qplot(speed, quality, data=viz, geom="point", color=factor(web), size = I(4)) + 
  coord_cartesian(x = c(0,11), y = c(0,11)) + theme_bw() +
  geom_text(aes(label = techs), color="black", vjust=1.4, hjust=1, size = I(4)) +
  theme(axis.text=element_blank(), axis.ticks=element_blank(), 
        legend.position = c(0.185, 0.1), legend.direction ="horizontal") +
  labs(color="Web Ready") + 
  scale_color_few() +
  labs(x = "Speed", y = "Quality") 
```

- Will this be shared on the web, in a document, or in a presentation? 
- Will the users want to interact with the output? 
- Is the visualization intended to be shared and used widely, or is it for a 
short exploration? 
- Do others need to be able to manipulate this image? 

\noindent Two main options. **Raster files** (.jpg, .png, and .gif) have a fixed scale, aspect ratio and size -- not good for resizing or reshaping. Easy to share.  
  
**Vector files** (.pdf and .svg) are zoomable and adjust to scale/aspect ratio changes. Viewable in browesrs, but not able to be opened in PowerPoint, Word, etc. \marginnote{If sharing matters most, go raster, if quality matters most consider vector files.}  
  

\noindent \emph{Learn More}

[www.jaredknowles.com](http://www.jaredknowles.com/presentations) • [http://forums.datacope.org](http://forums.datacope.org) • [http://eagereyes.org/about](http://eagereyes.org/about) • [http://datastori.es/](http://datastori.es/)

[^tufte]: http://www.edwardtufte.com/tufte/books_be
[^jayz]: (adapted from Jay-Z, 2003)