---
title: "Data Visualization Fieldguide for Education"
author: "Jared E. Knowles"
date: "March 2015"
output: 
  rmarkdown::tufte_handout:
    keep_tex: true
---

```{r knitrsetup,echo=FALSE,eval=TRUE,results="hide",error=FALSE,message=FALSE,warning=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, echo=FALSE, comment=NA)
library(eeptools)
library(ggplot2)
library(ggthemes)
require('hexbin')
source("data/simulate_data.R")
# load("data/cache.rda")
load("data/smalldata.rda")
```

Edward Tufte reminds you: [^tufte]

- Have a properly chosen format and design
- Use words, numbers and drawing together
- Reflect balance, proportion, and a sense of relevant scale
- Display an accessible complexity of detail 
- Have a anarrative quality, tell a story about the data
```{r, fig.margin = TRUE, fig.cap = "Sometimes the story is very simple."}
library(plyr)
plotdf <- ddply(iris, .(Species), summarize, Plength = mean(Petal.Length))
qplot(Species, Plength, data = plotdf, width = I(.5),
       geom='bar', 
      stat = 'identity') + theme_tufte() + labs(y = "Petal Length (cm)") 
```
- Draw in a professional manner, with the technical details done with care
- Avoid content-free decoration, including **chartjunk**

## Planning

Data visualization is a tool for communicating **a 
specific feature** of a dataset in an approachable and efficient manner.\marginnote{
Data visualization is done in service of the audience. If the goal is not to convey 
information, it is art!}

- Who is the audience? What is their background? Their biases?
- What is the context? What would the user consider to be *good news*, what would 
be *bad news*? How different are these? 
- What does the design style say? 
```{r, fig.margin = TRUE, fig.cap = "Theme says a lot. This says: `I am in a hurry'."}
qplot(Sepal.Length, Petal.Length, data = iris, color = Species) + theme_excel()
```
- Are you designing this visualization for a specific data set, or to display 
different data in a dashboard environment? 
- How long will this visualization last? Will it be updated next year?
- Does it need to be interactive?

```{r, fig.fullwidth = TRUE, fig.width = 10, fig.height = 2, fig.cap = "Small multiples spread the data out. But be careful, can the user draw meaningful distinctions between the groups from the data alone?"}
qplot(attday, readSS, data = df) + theme_bw() + 
  facet_wrap(~race, nrow=1) + labs(x = "Attendance Days", y = "Reading Score")
```

How you turn dimensions in the into visual cues for your audience is everything.

## Dimensionality 

*If you got big data, I feel sorry for you son. I got 99 problems but 
data viz ain't one.* [^jayz]

**Small Multiples**
  * Allow the viewer to draw comparisons across small plots

**Reduce the ink**
  * Each point has too much "weight", so make them weigh less
  * This has another advantage of depicting uncertainty through less depth, useful 
  when conveying data from a statistical model or sampling strategy
  
```{r, fig.margin = TRUE, fig.cap = "Use the alpha channel - it is intuitive."}
qplot(attday, readSS, data = df, alpha = I(0.5)) + theme_bw() + labs(x = "Attendance Days", y = "Reading Scale Score")
```

**Add Reference Points**
  * Big data is overwhelming without signposts, so give plenty of them to your viewer.

```{r, fig.margin = TRUE, fig.cap = "Scale invites the user to make comparisons."}
qplot(attday, readSS, data = df[df$race=="H" | df$race=="W", ], 
      alpha = I(0.5)) +   theme_bw() + 
  geom_hline(yintercept = 600, size = I(1.02), color = 'dark green') + 
#   geom_hline(yintercept = 530, , size = I(1.02), 
#              linetype = 2, color = I("dark blue")) +
#   geom_hline(yintercept = 480, , size = I(1.02), 
#              linetype = 2, color = I("goldenrod")) +
  geom_hline(yintercept = 420, , size = I(1.02), 
             linetype = 2, color = I("red")) + 
  annotate(geom="text", x = 135, y = 630, label = "Advanced", size = 3) + 
#   annotate(geom="text", x = 135, y = 560, label = "Proficient", size = 3) + 
#   annotate(geom="text", x = 135, y = 510, label = "Basic", size = 3) + 
  annotate(geom="text", x = 135, y = 450, label = "Minimal", size = 3) + 
      facet_wrap(~race, nrow=2) + 
  labs(x = "Attendance Days", y = "Reading Scale Score") 
```

- Model the data

```{r fig.margin = TRUE, fig.cap = "But why invite, when you can make comparisons directly and show those instead? This figure shows only the loess curves of the relationship between attendance and reading scores."}
ggplot(student_long[student_long$grade==5,],
       aes(y = readSS, x = attday, color = race))+
  geom_smooth(size=2) + theme_bw() + 
  scale_colour_few() + 
  labs(x="Attendance", y="Reading Score") + 
  scale_y_continuous(limits=c(420, 580)) + labs(colour = "Race") +
  theme(legend.position = c(0.35, 0.18), legend.direction = "horizontal")

```

# Review

**Before you submit, check that you have:**

- **Axis labels** and a **title**: Is your purpose clear?
- A **legend**: Are all symbols explained?
- A **scale**: Is the relation of data points clear?
- **Annotations** Is there enough context? Are key points highlighted?

Review this with *fresh eyes*. Can someone who has not seen this data before 
undertand and interpret the main story here?

Remember: If a picture is worth a thousand words, a good data visualization must 
always be *better than a table.*

## Technical Details

**The final format and use case should inform the design:**

```{r technologies,echo=FALSE,fig.margin = TRUE, fig.cap = "An opinonated ordering of the tools available to do data viz. Not comprehensive or complete and your mileage will vary depending on your comfort with certain technologies."}
techs<-c("Adobe", "R", "Stata/SPSS", "D3", "Tableau", "JavaScript", 
         "Processing", "Excel", "SAS")
speed  <- c(1.5, 7, 5, 6, 9, 4, 5, 9, 4)
quality <- c(10, 9, 5, 9, 8, 8, 9, 2, 4)
web <- c("Yes", "Yes", "No", "Yes", "No", "Yes", "Yes", "No", "No")
viz <- data.frame(techs=techs, speed=speed, quality=quality, web=web)

qplot(speed, quality, data=viz, geom="point", color=factor(web), size = I(4)) + 
  coord_cartesian(x = c(0,11), y = c(0,11)) + theme_bw() +
  geom_text(aes(label = techs), color="black", vjust=1.4, hjust=1, size = I(4)) +
  theme(axis.text=element_blank(), axis.ticks=element_blank(), 
        legend.position = c(0.185, 0.1), legend.direction ="horizontal") +
  labs(color="Web Ready") + 
  scale_color_few() +
  labs(x = "Speed", y = "Quality") 
```

**Raster Files** 

- **jpg** , **png** , **gif**. 
- Fixed scale, aspect ratio, and size
- Reasonable file size
- Get blurry when you blow them up or stretch them
- Viewable in almost any image viewing and editing system, including any modern web browser, PowerPoint, etc.

**Vector Files**

- **pdf** and **svg**
- Infinitely zoomable, adjustable on the fly
- Larger file size
- Viewable and usable in fewer systems. SVGs can be used in modern web browsers. PDFs included in other PDF reports. 

[^tufte]: http://www.edwardtufte.com/tufte/books_be
[^jayz]: (adapted from Jay-Z, 2003)